Sperm do not evolve to collaborate in female meiotic drive.

### 2014-04-23 writeup_and_notes/male_dependent_drive.tex
Added some analytical results and fixed a few typos

Despite the apparent unity of the organism, alleles can
gain an evolutionary advantage at a cost to  individual fitness
\citep{Burt2006}, often by exploiting meiosis and gametogenesis.
Because only one of the four products of female meiosis is transmitted to the egg, female meiosis is particularly vulnerable to such exploitation \citep{Sandler1957,Pardo-ManuelDeVillena2001a}. 
An allele that biases female meiosis in its favor (i.e. a meiotic driver), may increase in frequency even if this driver entails a pleiotropic fitness cost \citep{Prout1973}, generating a genetic conflict between the success of the driver and organismal fitness.
Meiotic drivers observed in nature (in both plants
\citep{Buckler1999,Fishman2005,Fishman2008}, and animals
\citep{Agulnik1990,Wu2005,Pardo-ManuelDeVillena2001c}) highlight this
conflict -- the selfish benefits of drive and the associated
pleiotropic fitness costs sustain a balanced polymorphism
\citep{Prout1973}, 
and often generate ongoing evolutionary escalations of drive suppressors and enhancers \citep{Dawe1996,Fishman2008}. 
The threat of meiotic drive to organismal fitness is potentially so
severe that many basic properties of
meiosis and o\"{o}genesis, including the initial genome doubling in
meiosis I \citep{Haig1991}, arrested female meiosis
\citep{Mira1998}, centromere machinery \citep{Malik2002a,Malik2009,Malik2009}, and sex differences in the recombination rate \citep{Haig2010,Brandvain2012} 
have perhaps evolved to enforce fairness by disrupting meiotic drive \citep{Rice2013}.

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is competed only after fertilization \citep{Masui_book}, 
creating ample opportunity for interaction between the sperm and female meiotic machinery. 
For example, an allele in sperm that facilitates meiotic drive by a genetically equivalent allele in female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through the population. 
At first sight, it seems as although female meiosis is primed to be exploited by selfish sperm systems.  
Why then is the requirement of fertilization to complete female meiosis so ubiquitous?

@@

It is also not the case that sperm is mechanistically incapable of influencing female meiosis.
In \emph{C. elegans} premature deployment of the aster (a vital component of mitotic
machinery) provided by the sperm can disrupts MII meiotic segregation
in the egg, leading to a triploid zygote \citep{McNally2012}. 
Thus, there is direct mechanistic evidence for that sperm can influence female meiosis.
 Additionally, genetic evidence from the In and Om  drive system sin mice suggests that the 
 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 Both of these female drive systems operate by distorting the second meiotic division, 
 and in both systems the outcome of female meiosis depends the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
While such genetic evidence is suggestive, we note the difficulty in
ruling out the alternative hypothesis of early selection on zygotes \citep[pages
52-54][]{Burt2006}.

%However,  we acknowledge the difficulty in ruling out early mortality
%as an alternative explanation of the 
%specifics of these putative drivers \citep[Give page numbers to
%discussion of this in][]{Burt2006}. \gc{should be careful here as the
 % Om locus is pretty convincingly a meiotic driver, doubts have just
 % been raised about whether IN is. Whether the sperm dep nature of
 % either is robust to this is I think more at question.}

@@

We first consider a driving allele neutral in heterozygotes, 
	but deleterious in homozygotes (whose relative fitness is
        $1-s$). This model is laid out in eqn. \ref{eqn.A1}-\ref{eqn.A3}.
In this model, a standard meiotic driver can always invade because 
	when rare it occurs predominantly in heterozygotes and it therefore drives without a fitness cost. 
However, a severe homozygous fitness cost ($s>XXd$) prevents  
	drivers from fixing  \citep{Prout1973}, 
	providing an opportunity for the evolution of
	drive suppressors (CITE - citep{XX}), corresponding well to 
	empirical examples of female meiotic drive.

In contrast to a traditional driver, which drives but pays effectively
no fitness cost as a rare deleterious recessive, 
	a self-promoting driver creates homozygotes
        when it drives, and therefore there is a drive-associated fitness cost even when rare. 
The self-promoting driver must overcome this homozygous fitness cost simply to spread when rare.
Therefore with a recessive fitness cost,
	the conditions allowing the invasion of a self-promoting driver
 	are far more restrictive than those for a standard meiotic driver (see Figure \ref{Invasion_space}).

When a deleterious recessive self-promoting driver does spread, 
	it spends much time at low frequency, because when rare it is unlikely to meet a complimentary sperm. 
However, once relatively common, a self-promoting driver spreads rapidly to fixation due to its
	positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}).  
This lack of a balanced polymorphism ( Figure \ref{Invasion_space}) precludes the evolution of an
allele that suppresses this form of meiotic drive in females.
\yb{POINT TO SOME MATHS OR FIGURE}

@@

>>>>

Despite the apparent unity of the organism, alleles can
gain an evolutionary advantage at a cost to  individual fitness
\citep{Burt2006}, often by exploiting meiosis and gametogenesis.
Because only one of the four products of female meiosis is transmitted to the egg, female meiosis is particularly vulnerable to such exploitation \citep{Sandler1957,Pardo-ManuelDeVillena2001a}. 
An allele that biases female meiosis in its favor (i.e. a meiotic driver), may increase in frequency even if this driver entails a pleiotropic fitness cost \citep{Prout1973}, generating a genetic conflict between the success of the driver and organismal fitness.
Meiotic drivers observed in both plants
\citep{Buckler1999,Fishman2005,Fishman2008}, and animals
\citep{Agulnik1990,Wu2005,Pardo-ManuelDeVillena2001c} highlight this
conflict -- the selfish benefits of drive and the associated
pleiotropic fitness costs sustain a balanced polymorphism
\citep{Prout1973}, 
and often generate ongoing evolutionary escalations of drive suppressors and enhancers \citep{Dawe1996,Fishman2008}. 
The threat of meiotic drive to organismal fitness is potentially so
severe that many basic properties of
meiosis and o\"{o}genesis, including the initial genome doubling in
meiosis I \citep{Haig1991}, arrested female meiosis
\citep{Mira1998}, centromere machinery \citep{Malik2002a,Malik2009,Malik2009}, and sex differences in the recombination rate \citep{Haig2010,Brandvain2012} 
have perhaps evolved to enforce fairness by disrupting meiotic drive \citep{Rice2013}.

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is competed only after fertilization \citep{Masui_book}, 
creating ample opportunity for interaction between the sperm and female meiotic machinery. 
Therefore, an allele in sperm that facilitates meiotic drive by a genetically equivalent allele in female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through the population. 
At first sight, female meiosis appears primed for exploitation by selfish sperm systems.  
Why then is the requirement of fertilization to complete female meiosis so ubiquitous?

@@

It is also not the case that sperm is mechanistically incapable of influencing female meiosis.
In \emph{C. elegans} premature deployment of the aster, a vital component of mitotic
machinery, provided by the sperm disrupts MII meiotic segregation
in the egg, leading to a triploid zygote \citep{McNally2012}. 
Thus, there is direct mechanistic evidence that sperm can influence female meiosis.
 Additionally, genetic evidence from the \emph{In} and \emph{Om}  drive systems in mice suggests that the 
 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes \citep[pages
52-54][]{Burt2006} in these cases is challenging, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
% Both of these female drive systems operate by distorting the second meiotic division, 
% and in both systems the outcome of female meiosis depends the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
%While such genetic evidence is suggestive, we note the difficulty in
%ruling out the alternative hypothesis of early selection on zygotes \citep[pages
%52-54][]{Burt2006}.

@@

We first consider a driving allele neutral in heterozygotes, 
	but deleterious in homozygotes (whose relative fitness is
        $1-s$). This model is laid out in eqn. \ref{eqn.A1}-\ref{eqn.A3}.
In this model, a standard meiotic driver can always invade because 
	when rare it occurs predominantly in heterozygotes and it therefore drives without a fitness cost. 
However, a severe homozygous fitness cost prevents  
	drivers from fixing  \citep{Prout1973}, 
	providing an opportunity for the evolution of
	drive suppressors (CITE - citep{XX}), corresponding well to 
	empirical examples of female meiotic drive.

In contrast to a traditional driver, which drives but pays effectively
no fitness cost as a rare deleterious recessive, 
	a self-promoting driver creates homozygotes
        when it drives, and therefore there is a drive-associated fitness cost even when rare. 
The self-promoting driver must overcome this homozygous fitness cost simply to spread when rare.
Therefore with a recessive fitness cost,
	the conditions allowing the invasion of a self-promoting driver
 	are far more restrictive than those for a standard meiotic driver (see Figure \ref{Invasion_space}).
\yb{When fully recessive, a self-promoting driver can only invade when $s$, the selection against the drive homozygote, is less than $\frac{2 d - 1}{4 d}$ (The tan region in Figure  \ref{Invasion_space}).}

When a deleterious recessive self-promoting driver does spread, 
	it spends much time at low frequency, because when rare it is unlikely to meet a complimentary sperm. 
However, once relatively common, a self-promoting driver spreads rapidly to fixation due to its
	positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}).  
This lack of a balanced polymorphism ( Figure \ref{Invasion_space}) precludes the evolution of an
allele that suppresses this form of meiotic drive in females.
\yb{In fact, while a rare self promoting driving can only invade if $s<\frac{2 d - 1}{4 d}$, a common self-promoting driver can fix over a broader range of parameter space (i.e. if $s<\frac{2d-1}{2}$). 
This system therefore displays bistability, when $\frac{2 d - 1}{4 d}<s<\frac{2d-1}{2}$, the driver will deterministically fix when introduced at a frequency greater than $\frac{1-2d+4ds}{2s(2d-1)}$, but will be lost otherwise.}

@@

\yb{\paragraph{Recessive fitness cost of self-promoting driver}
When the self-promoting driver has a recessive fitness cost, $s$ genotype frequencies after selection and drive equal
\begin{equation}	x'_{BB} =\frac{x_{B}^2(1-s)}{\bar{w} }  \Big( F+(1-F)(x_{B}+2x_{A} d) \Big) 	\end{equation}
\begin{equation}	x'_{AA} =\frac{1}{\bar{w} }  \Big(x_{A}^2\Big) 	\end{equation}
and
\begin{equation}	x'_{AB} =\frac{2 x_{A}x_{B} }{\bar{w} }  \Big(1-x_{B}(d-1/2)(1-F) \Big)	\end{equation}

its change in frequency across generations $\Delta f_B$ is
\begin{equation}
\frac{f_A f_B^2}{\bar{w}} \left( [1-F][ d(1-2 f_A s) - f_B s-1/2] - s  F \right)
\end{equation}
where $F$ is the deviation from Hardy-Weinberg expectations, and $f_A$ is the frequency of the non-driving allele. 
For analytical solutions presented above, we assume that $F$ 
}
